Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy

### 2025-06-03 Text/Articles/RENYI_ENTROPY_SAR_2025/R1-Renyi-entropy-sar.qmd
Add Suggested changes by Prof. ADCN

\textcolor{blue}{
Recent studies highlight the role of entropy in SAR classification. 
Cassetti et al. [@Cassetti2022] evaluated entropy estimators in both supervised and unsupervised models. 
Gallet et al. [@Gallet2024] proposed a Rényi divergence-based framework for explainable classification. 
Parikh et al. [@Parikh2019] discussed challenges in deep learning for SAR, including limited availability of labeled data and the complexity of tuning hyperparameters. These works motivate lightweight, interpretable alternatives.
}

@@

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which can describe roughness&nbsp;[@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{equation}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    = \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gamma1}
\end{equation}
and
\begin{multline}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    = \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}\\
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{multline}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
The $\Gamma_{\text{SAR}}$  model is a particular case of the $\mathcal{G}^0_I$ distribution&nbsp;[@Frery1997]: for $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) \text{ when } \alpha\to-\infty.
$$

@@

<!-- Introduced by Alfréd Rényi in 1961&nbsp;[@renyi1961measures], this measure generalizes several well-known entropies, including Shannon's&nbsp;[@Ribeiro2021].  -->
For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda \in \mathbbm R_+ \setminus \{1\}$, is defined as:
\begin{equation}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{equation}
Using&nbsp;\eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and the $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L)  \\ + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr],
\end{multline}
and
\begin{multline}  
\label{eq-HGI0}  
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr) = H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr) \;\;+ \\ \ln(-1 - \alpha)  
+ \frac{1}{1 - \lambda} \Bigl[ \lambda\bigl(\ln\Gamma(L - \alpha) - \ln\Gamma(-\alpha)\bigr) \\ 
+ \ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr) - \ln\Gamma\bigl(\lambda(L - \alpha)\bigr) \\
+ \bigl(\lambda(L-1)+1\bigr)\ln\lambda \Bigr].  
\end{multline}
<!-- The derivation of the latter result is proven in the Appendix. -->
Equation&nbsp;\eqref{eq-HGI0} can be read as:
$$
H_{\lambda}(\mathcal{G}^0_I)
=
\underbrace{H_\lambda\!\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{baseline entropy}}
\;\;\;\;\; +
\underbrace{\Delta_\alpha}_{\text{extra entropy caused by texture}},
$$
\textcolor{blue}{where the excess term $\Delta_\alpha$ depends on the roughness parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle) this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.}
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the coloured curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

Let $\bm{Z}=(Z_1, Z_2,\ldots,Z_n)$ be an independent and identically distributed random sample of size $n$ from a distribution $F$, and let $Z_{(1)} \leq Z_{(2)} \leq \dots \leq Z_{(n)}$ denote its order statistics.
The $m$-spacing density estimator is defined as:
$$
f_n(Z_{(i)}) = \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}},
$$
where $Z_{(i-m)} = Z_{(1)}$ when $i \leq m$, and $Z_{(i+m)} = Z_{(n)}$ if $i \geq n - m$. 
The coefficient $c_i$ is given by:
$$
c_i = 
\begin{cases}
\frac{m + i - 1}{m}, & \text{if } 1 \leq i \leq m, \\%[6pt]
2, & \text{if } m+1 \leq i \leq n - m, \\%[6pt]
\frac{n + m - i}{m}, & \text{if } n - m + 1 \leq i \leq n.
\end{cases}
$$
<!-- Following Vasicek&nbsp;[@vasicek1976test] and Ebrahimi et al.&nbsp;[@Ebrahimi1994] for Shannon entropy estimation, and using the $m$-spacing density estimator,  -->
The Rényi entropy can be estimated with this density estimator as
\begin{align}
\label{eq:est_R}
\widehat{H}_\lambda(\bm{Z}) = \frac{1}{1 - \lambda} \ln \left[\frac{1}{n} \sum_{i=1}^{n} \left( \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}} \right)^{\lambda - 1} \right].
\end{align}
This estimator is asymptotically consistent, i.e., it converges in
probability to the true value when $m,n\rightarrow\infty$ and
$m/n\rightarrow0$. 
We used the heuristic spacing $m=\left[\sqrt{n}+0.5\right]$.

@@

\textcolor{blue}{
We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator using simulated samples of size $n = 49$, drawn from the $\Gamma_{\text{SAR}}$ distribution.
We computed the bias and the mean squared error (MSE) for vaying $\lambda$ with a Monte Carlo experiment.
We found that for $L > 1$, $\lambda = 0.9$ consistently yields the lowest MSE while maintaining a low bias, thus offering a favorable trade-off between bias and variance. 
As shown in Fig. \ref{fig-optimal_order}, $\lambda = 0.85$ produces slightly lower bias, but leads to higher MSE, supporting the choice of $\lambda = 0.9$.}

@@

Following Refs.&nbsp;[@Frery2024;@Alpala2024], we refined the non-parametric entropy estimator $\widehat{H}_{\lambda}$ in&nbsp;\eqref{eq:est_R} with bootstrap, obtaining $\widetilde{H}_{\lambda}$:
\begin{equation*}
\widetilde{H}_{\lambda} = 2\widehat{H}_{\lambda}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{H}_{\lambda}(\bm{Z}^{(b)}),
\end{equation*}
where $B$ is the number of bootstrap replications, and $\bm{Z}^{(b)}$ is the $b$-th resampled dataset obtained by drawing $n$ observations with replacement from $\bm{Z}$.

A Monte Carlo study with $1000$ replications for each sample size $n \in \{9, 25, 49, 81, 121\}$ from the $\Gamma_{\text{SAR}}$ ($L=5,\ \mu=1$) confirms that for $\lambda=0.9$, the bootstrap-corrected estimator $\widetilde{H}_{\lambda}$ ($B=200$) reduces both bias and MSE compared to the original $\widehat{H}_{\lambda}$, with significant improvements for small sample sizes, as shown in Fig. \ref{fig-bias_mse}.
```{r Simulated_data_bias_B1, echo=FALSE, message=FALSE, cache = TRUE, autodep = TRUE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

## The Proposed Test
The test statistic quantifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0},
$$
\textcolor{blue}{the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids estimating parameters of the $\mathcal{G}^0_I$ model such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.}

Fig.&nbsp;\ref{fig-densities} shows the empirical distribution of $S_{\widetilde{H}_{\lambda}}$ under $\mathcal{H}_0$, obtained from $10^4$ Monte Carlo simulations with varying sample sizes ($n \in {49,81,121}$), $\lambda=0.9$, and $L \in {5,18}$. The concentration around zero confirms the test's expected behavior, while the heavy tails indicate sensitivity to deviations, enabling the detection of subtle texture changes.

@@

Vasicek&nbsp;[@vasicek1976test] proved that, for sufficiently large samples, $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$ asymptotically follows a normal distribution:
\begin{equation*}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) \overset{d}{\longrightarrow} \mathcal{N}(\mu_S,\,\sigma^{2}_S)\,,
\end{equation*}
where $d$ represents convergence in distribution. Here, $\mu_S  = \mathbb{E}[S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)]$ and $\sigma^{2}_S = \text{Var}[S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)]$ are the theoretical mean and variance of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$.

The standardized test statistic 
$\varepsilon = ({S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) - \widehat{\mu}_S})/{\widehat{\sigma}_S}$ is asymptotically standard normal distributed,
where $\widehat{\mu}_S$ and $\widehat{\sigma}_S$ are the empirical mean and standard deviation of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$, obtained by Monte Carlo simulations under the null hypothesis. 
Thus, we compute the $p$-values as $2\Phi(-|\varepsilon|)$,
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal distribution.
<!-- Fig.&nbsp;\ref{fig-density_standardized} shows smoothed histograms of standardized test statistics. -->

>>>>

Recent studies highlight the role of entropy in SAR classification. 
Cassetti et al. [@Cassetti2022] evaluated entropy estimators in both supervised and unsupervised models. 
Gallet et al. [@Gallet2024] proposed a Rényi divergence-based framework for explainable classification. 
Parikh et al. [@Parikh2019] discussed challenges in deep learning for SAR, including limited availability of labeled data and the complexity of tuning hyperparameters. These works motivate lightweight, interpretable alternatives.

@@

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which can describe roughness&nbsp;[@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{equation}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    = \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
\end{equation}
and
\begin{multline}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    = \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}\\
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{multline}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
The $\Gamma_{\text{SAR}}$  law is a particular case of the $\mathcal{G}^0_I$ distribution&nbsp;[@Frery1997]: for $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) \text{ when } \alpha\to-\infty.
$$

@@

<!-- Introduced by Alfréd Rényi in 1961&nbsp;[@renyi1961measures], this measure generalizes several well-known entropies, including Shannon's&nbsp;[@Ribeiro2021].  -->
For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda \in \mathbbm R_+ \setminus \{1\}$ is defined as:
\begin{equation}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{equation}
Using&nbsp;\eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and the $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L)  \\ + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr]
\end{multline}
and
\begin{multline}  
\label{eq-HGI0}  
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr) = H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr) \;\;+ \\ \ln(-1 - \alpha)  
+ \frac{1}{1 - \lambda} \Bigl[ \lambda\bigl(\ln\Gamma(L - \alpha) - \ln\Gamma(-\alpha)\bigr) \\ 
+ \ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr) - \ln\Gamma\bigl(\lambda(L - \alpha)\bigr) \\
+ \bigl(\lambda(L-1)+1\bigr)\ln\lambda \Bigr].  
\end{multline}
<!-- The derivation of the latter result is proven in the Appendix. -->
Equation&nbsp;\eqref{eq-HGI0} can be read as:
$$
H_{\lambda}(\mathcal{G}^0_I)
=
\underbrace{H_\lambda\!\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{baseline entropy}}
\hspace{1.8em} + \hspace{-1.0em}
\underbrace{\Delta_\alpha}_{\text{extra entropy caused by texture}}\hspace{-3.0em},
$$

\textcolor{blue}{where the excess term $\Delta_\alpha$ depends on the roughness parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.}
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the coloured curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

Let $\bm{Z}=(Z_1, Z_2,\ldots,Z_n)$ be an independent and identically distributed random sample of size $n$ from $Z \sim F$, and let $Z_{(1)} \leq Z_{(2)} \leq \dots \leq Z_{(n)}$ denote its order statistics.
The $m$-spacing density estimator is defined as:
$$
f_n(Z_{(i)}) = \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}},
$$
where $Z_{(i-m)} = Z_{(1)}$ when $i \leq m$, and $Z_{(i+m)} = Z_{(n)}$ if $i \geq n - m$. 
The coefficient $c_i$ is given by:
$$
c_i = 
\begin{cases}
\frac{m + i - 1}{m}, & \text{if } 1 \leq i \leq m, \\%[6pt]
2, & \text{if } m+1 \leq i \leq n - m, \\%[6pt]
\frac{n + m - i}{m}, & \text{if } n - m + 1 \leq i \leq n.
\end{cases}
$$
<!-- Following Vasicek&nbsp;[@vasicek1976test] and Ebrahimi et al.&nbsp;[@Ebrahimi1994] for Shannon entropy estimation, and using the $m$-spacing density estimator,  -->
The Rényi entropy can be estimated with this density estimator as
\begin{align}
\label{eq:est_R}
\widehat{H}_\lambda(\bm{Z}) = \frac{1}{1 - \lambda} \ln \left[\frac{1}{n} \sum_{i=1}^{n} \left( \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}} \right)^{\lambda - 1} \right].
\end{align}
This estimator is asymptotically consistent, i.e., it converges in
probability to the true value when $m,n\rightarrow\infty$ and
$m/n\rightarrow0$. 
We use the heuristic spacing $m=\left[\sqrt{n}+0.5\right]$.

@@

\textcolor{blue}{
We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator using simulated samples of size $n = 49$, drawn from  $Z\sim \Gamma_{\text{SAR}}(5,1)$ distribution.
We computed the bias and the mean squared error (MSE) for vaying $\lambda$ with a Monte Carlo experiment.
We found that for $L > 1$, $\lambda = 0.9$  yields the lowest MSE while maintaining a low bias, thus offering a favorable trade-off between bias and variance. 
As shown in Fig. \ref{fig-optimal_order}, $\lambda = 0.85$ produces slightly lower bias, but leads to higher MSE, supporting the choice of $\lambda = 0.9$.}

@@

Following Refs.&nbsp;[@Frery2024;@Alpala2024], we refine the non-parametric entropy estimator $\widehat{H}_{\lambda}$ in&nbsp;\eqref{eq:est_R} with bootstrap, obtaining $\widetilde{H}_{\lambda}$:
\begin{equation*}
\widetilde{H}_{\lambda} = 2\widehat{H}_{\lambda}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{H}_{\lambda}(\bm{Z}^{(b)}),
\end{equation*}
where $B$ is the number of bootstrap replications, and $\bm{Z}^{(b)}$ is the $b$-th resampled dataset obtained by drawing $n$ observations with replacement from $\bm{Z}$.

A Monte Carlo study with $1000$ replications for each sample size $n \in \{9, 25, 49, 81, 121\}$ from the $\Gamma_{\text{SAR}}$ (5,1) confirms that for $\lambda=0.9$, the bootstrap-corrected estimator $\widetilde{H}_{\lambda}$ ($B=200$) reduces both bias and MSE compared to the original $\widehat{H}_{\lambda}$, with significant improvements for small sample sizes, as shown in Fig. \ref{fig-bias_mse}.
```{r Simulated_data_bias_B1, echo=FALSE, message=FALSE, cache = TRUE, autodep = TRUE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

## The Proposed Test
The test statistic quantifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0}\hspace{-0.5em},
$$
\textcolor{blue}{the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids estimating  $\mathcal{G}^0_I$ parameters  such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.}

Fig.&nbsp;\ref{fig-densities} shows the empirical distribution of $S_{\widetilde{H}_{\lambda}}$ under $\mathcal{H}_0$, obtained from $10^4$ Monte Carlo experiments with varying sample sizes ($n \in {49,81,121}$), $\lambda=0.9$, and $L \in {5,18}$. The concentration around zero confirms the expected behavior of the test, while the heavy tails indicate sensitivity to deviations, enabling the detection of subtle texture changes.

@@

Vasicek&nbsp;[@vasicek1976test] proved that, for sufficiently large samples, $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$ asymptotically follows a normal distribution:
\begin{equation*}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) 
\overset{\mathcal{D}}{\underset{n \to \infty}{\longrightarrow}} 
\mathcal{N}(\mu_S,\,\sigma^{2}_S),
\end{equation*}
where $\mathcal{D}$ represents convergence in distribution. Here, $\mu_S  = \mathbb{E}[S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)]$ and $\sigma^{2}_S = \text{Var}[S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)]$ are the theoretical mean and variance of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$.

The standardized test statistic 
$\varepsilon = ({S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) - \widehat{\mu}_S})/{\widehat{\sigma}_S}$ is asymptotically standard normal distributed,
where $\widehat{\mu}_S$ and $\widehat{\sigma}_S$ are the estimated mean and standard deviation of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$, obtained by Monte Carlo simulations under the null hypothesis. 
Thus, we compute the $p$-values as $2\Phi(-|\varepsilon|)$,
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal law.
<!-- Fig.&nbsp;\ref{fig-density_standardized} shows smoothed histograms of standardized test statistics. -->
